---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

**Preliminary work.**

Load in package dependencies.
```{r results='hide'}
library(plyr)
library(dplyr)
library(ggplot2)
```

Unzip the "activity.zip" file and save "activity.csv"" to a separate data folder in the working directory if the file/directory doesn't already exist.

```{r}
if (!file.exists("./data")) {
    dir.create("./data")
}
if (!file.exists("./data/activity.csv")) {
    unzip("activity.zip", files="activity.csv", exdir="./data")
}
```

Read in the data.
```{r}
activity <- read.csv("./data/activity.csv", header = TRUE, colClasses=c("numeric", "character", "numeric"))
dim(activity)
```

Activity data summary:
```{r}
summary(activity)
```

Convert the date column values from character to Date:
```{r}
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
```
## What is mean total number of steps taken per day?

**1. Calculate the total number of steps taken per day, ignoring the missing values in the dataset.**

```{r}
## dailysteps <- select(rowsum(activity, activity$date, na.rm = FALSE), steps)  
dailySteps <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
```

**2. Make a histogram of the total number of steps taken each day.**

```{r fig.height=6}

## hist(dailysteps$steps, xlab = "Number of Steps", labels = TRUE, main="Histogram of Daily Steps")
qplot(dailySteps, xlab = "Total Daily Steps", ylab = "Frequency")
```

**3. Calculate and report the mean and median of the total number of steps taken per day**

```{r echo=FALSE}
dailyMean <- mean(dailySteps)
dailyMedian <- median(dailySteps)
```
Mean: 

```{r}
mean(dailySteps)
```

Median: 

```{r}
median(dailySteps)
```

- The daily steps **mean** is `r dailyMean`.

- The daily steps **median** is `r dailyMedian`.

## What is the average daily activity pattern?

**1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**
```{r}
avgIntervalSteps <- aggregate(
                        list(stepsMean=activity$steps),
                        by = list(interval=activity$interval),
                        FUN = mean, 
                        na.rm = TRUE
                    )
```

Make a time series plot

```{r}
ggplot(avgIntervalSteps, aes(x=interval, y=stepsMean)) + 
    geom_line() + 
    xlab("5-Minute Interval") + 
    ylab("Average Number of Steps Taken")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
mostActiveInterval <- avgIntervalSteps[which.max(avgIntervalSteps$steps), c("interval")]

mostActiveInterval
```

**Answer:** On average, the most active 5-minute interval is **`r mostActiveInterval`**.

## Imputing missing values

**1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r}
numMissing <- nrow(activity[is.na(activity$steps),])
numMissing
```

**Answer:** The total number of missing steps is **`r numMissing`**.

**2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

```{r}
## Replace missing values with the mean for that 5-minute interval.

modActivity <- merge(activity, avgIntervalSteps, by = "interval", all.x = TRUE)
```

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r}
## Replace missing date with the 5-min interval, remove the extra data and sort by date, interval.

modActivity[is.na(modActivity$steps), c("steps")] <- modActivity[is.na(modActivity$steps), c("stepsMean")]
modActivity <- select(modActivity, steps, date, interval)
modActivity <- arrange(modActivity, date, interval)
modActivity$date <- as.Date(modActivity$date)

head(modActivity)
```

**4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.**

Mean and median daily steps are the same after replacing the NAs with the interval means.

```{r}
modDailySteps <- tapply(modActivity$steps, modActivity$date, sum)
qplot(modDailySteps, xlab = "Total Daily Steps", ylab = "Frequency")
```

```{r echo=FALSE}
modDailyMean <- mean(modDailySteps)
modDailyMedian <- median(modDailySteps)
```
Mean: 

```{r}
mean(modDailySteps)
```

Median: 

```{r}
median(modDailySteps)
```

What is the impact of imputing missing data on the estimates of the total daily number of steps?
**Answer:** The results show a more natural bell curve over the day; the spike at interval 0 is gone.

Do these values differ from the estimates from the first part of the assignment?

**Answer:** Mean and Median are higher than the values in the first part of the assignment. Unlike the first part of the assignment, the mean and median values are now equal.

- The daily steps **mean** is `r modDailyMean`.

- The daily steps **median** is `r modDailyMedian`.

## Are there differences in activity patterns between weekdays and weekends?

**Answer:** Yes, the activity patterns differ between weekdays and weekends.

**1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.**

```{r}
modActivity$daytype <- as.factor(ifelse ( weekdays(modActivity$date) %in% c("Saturday","Sunday"), "weekend", "weekday" ) )
```


**2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

```{r}
avgModActivity <- aggregate(
                    steps ~ interval + daytype,
                    data = modActivity,
                    FUN = mean
                  )

ggplot(avgModActivity, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(daytype ~ .) + 
    xlab("5-Minute Interval") + 
    ylab("Average Number of Steps Taken")
```